---
title: "Midterm 2"
author: "Srinidhi Viswanathan"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your code should be organized, clean, and run free from errors. Be sure to **add your name** to the author header above. You may use any resources to answer these questions (including each other), but you may not post questions to Open Stacks or external help sites. There are 10 total questions.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean! Your plots should use consistent aesthetics throughout. Feel free to be creative- there are many possible solutions to these questions!  

This exam is due by **12:00p on Tuesday, February 23**.  

## Load the libraries
```{r}
library(tidyverse)
library(janitor)
library(here)
library(skimr)
options(scipen=999) #disables scientific notation when printing
```

## Gapminder
For this assignment, we are going to use data from  [gapminder](https://www.gapminder.org/). Gapminder includes information about economics, population, social issues, and life expectancy from countries all over the world. We will use three data sets, so please load all three.  

One thing to note is that the data include years beyond 2021. These are projections based on modeling done by the gapminder organization. Start by importing the data.
```{r}
population <- read_csv("data/population_total.csv")
population
```

```{r}
income <- read_csv("data/income_per_person_gdppercapita_ppp_inflation_adjusted.csv")
income
```

```{r}
life_expectancy <- read_csv("data/life_expectancy_years.csv")
life_expectancy
```

1. (3 points) Once you have an idea of the structure of the data, please make each data set tidy and store them as new objects. You will need both the original and tidy data!
##### For the population data frame:
```{r}
skim(population)
```

```{r}
pop_tidy <- population %>% 
  pivot_longer(-country, #country does not pivot
               names_to = "year", 
               values_to = "population"
               )
pop_tidy
```

##### For the income data frame:
```{r}
skim(income)
```

```{r}
income_tidy <- income %>% 
  pivot_longer(-country, #country does not pivot
               names_to = "year", 
               values_to = "gdp_per_capita"
               )
income_tidy
```

##### For the life expectancy data frame:
```{r}
skim(life_expectancy)
```

```{r}
life_tidy <- life_expectancy %>% 
  pivot_longer(-country, #country does not pivot
               names_to = "year", 
               values_to = "life_expectancy"
               )
life_tidy
```
###### There are NAs in this data, represented by "NA"

2. (1 point) How many different countries are represented in the data? Provide the total number and their names. Since each data set includes different numbers of countries, you will need to do this for each one.
##### For the population data:
```{r}
pop_tidy %>% 
  group_by(country) %>% 
  summarize(n_country=n_distinct(country))
```
###### There are 195 different countries represented in the population data.

You could also think about something like this...
```{r}
n_distinct(pop_tidy$country)
```

##### For the income data:
```{r}
income_tidy %>% 
  group_by(country) %>% 
  summarize(n_country=n_distinct(country))
```
###### There are 193 different countries represented.

##### For the life expectancy data:
```{r}
life_tidy %>% 
  group_by(country) %>% 
  summarize(n_country=n_distinct(country))
```
###### There are 187 countries represented.

## Life Expectancy  

3. (2 points) Let's limit the data to the past 100 years (1920-2020). For these years, which country has the highest life expectancy? How about the lowest life expectancy?  

```{r}
hundred_life <- life_tidy %>% 
  filter(year<=2020, year>=1920) %>% 
  group_by(country) %>% 
  summarize(mean_life_expectancy=mean(life_expectancy, na.rm=T))
hundred_life
```


Highest
```{r}
hundred_life %>% 
  arrange(desc(mean_life_expectancy))
```
```{r}
life_tidy %>% 
  filter(country=="Andorra") %>% 
  filter(year>=1920, year<=2020)
```
```{r}
life_tidy %>% 
  filter(country=="Sweden") %>% 
  filter(year>=1920, year<=2020)
```

##### Andorra has the highest average life expectancy from the years (although, this data does need to be taken with a grain of salt because Andorra has a lot of NAs in its data. The data starts in 1970 and ends in 2019.)
##### Sweden, on the other hand, has the highest life expectancy of the data that has no NAs in this time frame.

Lowest
```{r}
hundred_life %>% 
  arrange(mean_life_expectancy)
```
```{r}
life_tidy %>% 
  filter(country=="Andorra") %>% 
  filter(year>=1920, year<=2020)
```
```{r}
life_tidy %>% 
  filter(country=="Mali") %>% 
  filter(year>=1920, year<=2020)
```
##### Again, the country that is shown to have the lowest life expectancy (the Central African Republic) has incomplete data.
##### The country that has lowest life expectancy with complete data is Mali.


4. (3 points) Although we can see which country has the highest life expectancy for the past 100 years, we don't know which countries have changed the most. What are the top 5 countries that have experienced the biggest improvement in life expectancy between 1920-2020?
```{r}
life_exp_improve <- life_tidy %>% 
  filter(year==2020|year==1920) %>% 
  group_by(country) %>% 
  summarize(improvement_years=max(life_expectancy)-min(life_expectancy))
life_exp_improve
```

```{r}
life_exp_improve %>% 
  arrange(desc(improvement_years))
```
##### Kuwait has had the biggest increase in life expectancy, followed by the Kyrgyz Republic, Terkmenistan, South Korea, and Tajikistan.

5. (3 points) Make a plot that shows the change over the past 100 years for the country with the biggest improvement in life expectancy. Be sure to add appropriate aesthetics to make the plot clean and clear. Once you have made the plot, do a little internet searching and see if you can discover what historical event may have contributed to this remarkable change.  
```{r}
life_tidy$year <- as.numeric(life_tidy$year)
```

```{r}
life_tidy %>% 
  filter(year<=2020,year>=1920) %>% 
  filter(country=="Kuwait") %>% 
  ggplot(aes(x=year, y=life_expectancy))+geom_point()+geom_line()+
  theme(text = element_text(size=6),
        axis.text.x = element_text(angle=60, hjust=1)) +
  theme(plot.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 10))+
  labs(title = "Life Expectancy in Kuwait (1920-2020)",
       x = "Year",
       y= "Life Expectancy")
  
```

##### In the 1930s, Kuwait's petroleum industry began. This led to an rapid increase in quality of life to those who lived in Kuwait in that era.

## Population Growth
6. (3 points) Which 5 countries have had the highest population growth over the past 100 years (1920-2020)?
```{r}
pop_growth_hundred <- pop_tidy %>% 
  filter(year==2020|year==1920) %>% 
  group_by(country) %>% 
  summarize(population_growth=max(population)-min(population)) %>% 
  arrange(desc(population_growth))
pop_growth_hundred
```
##### The five countries that had the most population growth were India, China, Indonesia, the United States, and Pakistan.

7. (4 points) Produce a plot that shows the 5 countries that have had the highest population growth over the past 100 years (1920-2020). Which countries appear to have had exponential growth? 
```{r}
pop_tidy$year <- as.numeric(pop_tidy$year)
```

```{r}
pop_tidy %>% 
  filter(year<=2020,year>=1920) %>% 
  filter(country=="India"|country=="China"|country=="Indonesia"|country=="United States"|country=="Pakistan") %>% 
  ggplot(aes(x=year, y=population, color=country))+geom_point(shape=1)+geom_line()+
  theme(text = element_text(size=6),
        axis.text.x = element_text(angle=60, hjust=1)) +
  theme(plot.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 10))+
  labs(title = "Population over Time",
       x = "Year",
       y= "Population")
```

##### China and India seem to have exponential growth. 


## Income
The units used for income are gross domestic product per person adjusted for differences in purchasing power in international dollars.

8. (4 points) As in the previous questions, which countries have experienced the biggest growth in per person GDP. Show this as a table and then plot the changes for the top 5 countries. With a bit of research, you should be able to explain the dramatic downturns of the wealthiest economies that occurred during the 1980's.
```{r}
gdp_increase_hundred <- income_tidy %>% 
  filter(year==2020|year==1920) %>% 
  group_by(country) %>% 
  summarize(gdp_growth=max(gdp_per_capita)-min(gdp_per_capita)) %>% 
  arrange(desc(gdp_growth))
gdp_increase_hundred
```
##### The five countries with the biggest growth of GDP are Qatar, Luxembourg, Singapore, Brunei, and Ireland.
```{r}
income_tidy$year <- as.numeric(income_tidy$year)
```

```{r}
income_tidy %>% 
  filter(year<=2020,year>=1920) %>% 
  filter(country=="Qatar"|country=="Luxembourg"|country=="Singapore"|country=="Brunei"|country=="Ireland") %>% 
  ggplot(aes(x=year, y=gdp_per_capita, color=country))+geom_point(shape=1)+geom_line()+
  theme(text = element_text(size=6),
        axis.text.x = element_text(angle=60, hjust=1)) +
 theme(plot.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 10))+
  labs(title = "GDP per capita over Time",
       x = "Year",
       y= "GDP per capita")
```

##### In the 1980s, there was a sever economic recession that was caused by the 1979 energy crisis.  This happened because oil prices increased dramatically due to the Iranian revolution.

9. (3 points) Create three new objects that restrict each data set (life expectancy, population, income) to the years 1920-2020. Hint: I suggest doing this with the long form of your data. Once this is done, merge all three data sets using the code I provide below. You may need to adjust the code depending on how you have named your objects. I called mine `life_expectancy_100`, `population_100`, and `income_100`. For some of you, learning these `joins` will be important for your project.  

life_expectancy_100
```{r}
life_expectancy_100 <- life_tidy %>% 
  filter(year<=2020,year>=1920)
  
```

population_100
```{r}
population_100 <- pop_tidy %>% 
  filter(year<=2020,year>=1920)
```

income_100
```{r}
income_100 <- income_tidy %>% 
  filter(year<=2020,year>=1920)
```

```{r}
gapminder_join <- inner_join(life_expectancy_100, population_100, by= c("country", "year"))
gapminder_join <- inner_join(gapminder_join, income_100, by= c("country", "year"))
gapminder_join
```

10. (4 points) Use the joined data to perform an analysis of your choice. The analysis should include a comparison between two or more of the variables `life_expectancy`, `population`, or `income.`
##### As the life expectancy of the current 5 most populous nations has gotten bigger, has the population also grown?
###### First, I'm going to find the current 5 most populous countries.
```{r}
gapminder_join %>% 
  filter(year==2020) %>% 
  arrange(desc(population))
```
###### The current 5 most populated countries are: China, India, the United States, Indonesia, and Pakistan.

###### Now, I can make a line plot comparing these 5 countries in terms of how their life expectancy has affected their population.
```{r}
gapminder_join %>% 
  filter(country=="China"|country=="India"|country=="United States"| country=="Indonesia"|country=="Pakistan") %>% 
  ggplot(aes(x=life_expectancy, y=population, color=country))+geom_point()+geom_line()+
 theme(plot.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 10))+
  labs(title = "Life Expectancy vs. Population for the 5 Most Populated Countries",
       x = "Life Expectancy",
       y= "Population")
```

##### The does seem to be a positive correlation between an increase in life expectancy and an increase in population. This is especially seen in China and India. In Indonesia, Pakistan, and the United States, the correlation is there, but is less pronounced.

